Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-22 Response_referees.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

List of major changes apported to the paper:
\begin{itemize}
 \item Much efforts have been spent in trying to clarify better the relation between localization properties and features of the computational basis needed 
 to provide precise results. This topic has been discussed in all the text. Major modifications under this perspective have been done in the introduction, in section II at pag. 5 before the analysis of fig. 2,in section IV and in the conclusions. 
 \item The introduction has been deeply revised. In particular we have reduced the length of the first two paragraph concerning the differences between accuracy and precision.
 \item We added a comment after eq. (7) on the non-hermiticity of the response density and corrected the equation (11).
 \item Added a dedicated subsection (actually the IIA at page 4)  to discuss the localization properties of the FS states for static perturbations. 
 \item Added a panel in figure 2 that contain the plot of the real part of $\alpha(\omega)$ in calculated in different computational domain. This new plot aims to evidence that the localization
 properties of th FS affect, on equal footing, both the real and the imaginary part of the linear response functional.
 \item We have reorganized the material after section III A. In particular we have introduced a new section (number IV in the actual version of the paper) in which we have discussed both the physical features of the excitation spectrum and the relative importance of localized vs. delocalized sector in expressing the susceptibility functional. This section also collects computational results for molecules beyond the case of $CO$ and benzene, already discussed in the previous version of the paper.  
 \item State something about the inclusion of other molecules...
\end{itemize}

@@

\begin{enumerate}

 \item \emph{On the localized sector of excitation. Dependence on the symmetry and dimension of the molecule.}

 In section IIIA we have provided formal arguments that evidence that the excitation spectrum splits into two sectors with distinct features. The first one
 is composed by a finite set of localized excitations with discrete eigenvalues, whereas the second one contains a continuum spectrum of delocalized objects.
 For excitation energies lower than the ionization potential (IP) of the molecule all the excitations belong to the first sector. In the intermediate energy range which goes from IP to the absolute value of the energy of the deepest occupied orbital the two sectors compenetrate so that there are localized excitation embedded in a continuum of delocalized ones, and finally only elements of the second sector exist for higher values of the energy.

 These formal results have been supported by numerical calculations in which the excitation spectrum of several molecules has been explicitly computed.
 The attribution of each excitation of the appropriate sector passes through the evaluation of the norm of the states $\phi_p^a$ and $\chi_p^a$ that build
 the dyadic representation of the excitation (see eq. 17). Only states with norm greater than a given tolerance value are considered and the associated 
 excitation is attributed to the first sector if the condition $\Omega_a < |\epsilon_p|$ with $\epsilon_a = \mathrm{max}\left(\epsilon_p\right)_{p\in a}$.
 This computational procedure provides results in optimum agreement with the formal arguments reported above. Indeed the excitation landscape of 
 (for instace the panel a o fig. 5 for the case of Benzene) evidences the presence of discrete excitations (the black bullets) embedded in a pseudo-continuum 
 represented by the blue filled square. Moreover the comparison between the excitation DOS computed in different computational domain shows that this way of labeling the excitation is correct since the elements attributed to the localized sector have energies that easily converge with respect to the increase of the simulation box, as expected.

 In the previous version of the paper this type of analysis was limited to the case of $CO$ and Benzene molecules. However the same structure of the excitation spectrum is quite general and turns out not to be affected by the size or symmetry of the molecule. To check this statement we have repeated
 the same computational analysis for several other molecules of different shape and dimension like the Fullerene and the Aflatoxin. Results for this studies 
 are presented as supplementary materials. Looking at the excitation landscape for these cases (extracted with the same procedure and tolerance parameters?) it...
 to be completed after the inclusion of the table.....

@@

\item \emph{Regarding the localized excitations which can still be found above the IP threshold (Fig 5a), is there any scope to target these, to the
exclusion of other states, so that a set of just the localized excitations can be calculated at reduced computational cost compared to the whole set?}

 During the review process we have deeply revised the final part of the paper (in particular the section IVB in the actual version of the manuscript) in order to clarify which is the relevance the two excitation sectors for obtaining an effective representation of the susceptibility functional.

 For a sake of concreteness this analysis has been declined for the relevant case of the computation of the dynamical polarizability. It emerges that the localized
 sector of excitation is sufficient to express the imaginary part of $\alpha(\omega)$ \emph{only} for values of $\omega$ below IP (see eq. 21). The evaluation of
 other quantities, namely $\mathrm{Im}(\alpha(\omega))$ above IP or $\mathrm{Re}(\alpha(\omega))$ in all the $\omega$ range requires a relevant contribution from
 both the sectors of excitation spectrum. So, to answer to the question raised by the referee we can state that localized excitations above IP can be isolated and computed using the procedure described in the previous point of this report, but their knowledge alone is not useful to extract any physical quantities.
 We can state that the multi-threshold structure of the excitation spectrum is surely interesting but is not very relevant for computational purposes.

### 2018-11-22 AnalyticStructure_main.tex
paper review and update referee report

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, 
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} 
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in ``weak'' 
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned 
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design} 
a localized behaviour of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has 
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ . 
This fact implies that the degrees of completeness of the computational basis should therefore be increased by adding the necessary degrees
of freedom for describe the FS in all the relevant domain. Stated otherwise, in order to achieve precise results in in this regime the 
underlying basis set has to able to express delocalized states.
%The degrees of completeness of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to 
%represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalized states.

@@

To further address this point we exploit the formal relation between response density and susceptibility functional %established in ~\eqref{SusceptibilityDef1} 
$\dm'(\omega) = \sop \chi(\omega)\op\Phi(\omega)$ and make usage of the diadic representation ~\eqref{rhoPrimeFluctuationStateDef1} of the response density. Pursuing this procedure provides the excitation-based
representation of FS as follows:
% Let us now consider the features of the response density operator of Eq.~\eqref{rhoPrimeFluctuationStateDef1}.
% The fluctuation state in this case can be written as
% \be\lb{fsinexc}
% \ket{f_p^{\mathbf r_j}(\omega)}=
% \sop \chi(\omega) \op{\mathbf r_j} \ket{\psi_p}
% = \sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
% \ket{w_p^a}\;,
% \ee
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;,
\ee
where we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
%The above equation shows that for \emph{any} value of $\omega$ both sectors of localized and delocalized states $\ket{w_p^a}$ contribute.
The above equation shows that both the sectors of localized and delocalized states $\ket{w_p^a}$ contribute to the FS for \emph{any} value of $\omega$.
However the discussion of Sec.~\ref{FluctuationState} has shown that the FS is a genuinely localized state below the IP threshold.
The equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
% Such a locality is evident for the imaginary part of the FS, which can be straightforwardly restricted to the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). For the real part of FS the \emph{entire} spectrum
% of excitations contribute.
The localization below threshold is manifest for the imaginary part of the FS, that is explicitly written as a the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). Instead, for the real part of FS, the \emph{entire} excitation spectrum contributes to the construction of the FS and the localization
\emph{has} to realize due to a non trivial cancellation of the unbound parts of the $\ket{w_p^a}$'s.

@@

The projection $\op O_\perp$ is therefore the relevant part for the commutator
\begin{equation}
\commutator{\dmnot}{\op O} = \commutator{\dmnot}{\op O_\perp} =
\dmnot \op O \op Q_0  - \op Q_0 \op O \dmnot  \;,
\end{equation}
consequently the Liouvillian superoperator, which contains by definition a commutator with $\hnot$ and $\dmnot$ in its bare and coupling part, is constructed such as $\Liouv \op O = \Liouv O_\perp$. In addition, the image operator satisfies the transverse condition
$\left( \Liouv \op O \right)_\parallel =0$.

>>>>

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, 
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} 
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in ``weak'' 
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned 
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design} 
a localized behaviour of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has 
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ . 
This fact implies that the degrees of completeness of the computational basis should therefore be increased by adding the necessary degrees
of freedom for describe the FS in all the relevant domain. Stated otherwise, to achieve precise results in this regime the 
underlying basis set has to able to express delocalized states.
%The degrees of completeness of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to 
%represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalized states.

@@

To further address this point we exploit the formal relation between response density and susceptibility functional %established in ~\eqref{SusceptibilityDef1} 
$\dm'(\omega) = \sop \chi(\omega)\op\Phi(\omega)$ and make usage of the dyadic representation ~\eqref{rhoPrimeFluctuationStateDef1} of the response density. Pursuing this procedure provides the excitation-based
representation of FS as follows:
% Let us now consider the features of the response density operator of Eq.~\eqref{rhoPrimeFluctuationStateDef1}.
% The fluctuation state in this case can be written as
% \be\lb{fsinexc}
% \ket{f_p^{\mathbf r_j}(\omega)}=
% \sop \chi(\omega) \op{\mathbf r_j} \ket{\psi_p}
% = \sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
% \ket{w_p^a}\;,
% \ee
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;,
\ee
where we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
%The above equation shows that for \emph{any} value of $\omega$ both sectors of localized and delocalized states $\ket{w_p^a}$ contribute.
The above equation shows that both the sectors of localized and delocalized states $\ket{w_p^a}$ contribute to the FS for \emph{any} value of $\omega$.
However the discussion of Sec.~\ref{FluctuationState} has shown that the FS is a genuinely localized state below the IP threshold.
The equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
% Such a locality is evident for the imaginary part of the FS, which can be straightforwardly restricted to the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). For the real part of FS the \emph{entire} spectrum
% of excitations contribute.
The localization below threshold is manifest for the imaginary part of the FS, that is explicitly written as a the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). Instead, for the real part of FS, the \emph{entire} excitation spectrum contributes to the construction of the FS and the localization
\emph{has} to realize due to a non trivial cancellation of the unbound parts of the $\ket{w_p^a}$'s.

@@

The projection $\op O_\perp$ is therefore the relevant part for the commutator
\begin{equation}
\commutator{\dmnot}{\op O} = \commutator{\dmnot}{\op O_\perp} =
\dmnot \op O \op Q_0  - \op Q_0 \op O \dmnot  \;,
\end{equation}
consequently the Liouvillian superoperator, which contains by definition a commutator with $\hnot$ and $\dmnot$ in its unperturbed and coupling part respectively,
is constructed such as $\Liouv \op O = \Liouv O_\perp$. In addition, the image operator satisfies the transverse condition
$\left( \Liouv \op O \right)_\parallel =0$.
